---
title: "Project 1: Data combination and summarization through functions"
author: Laraib Azmat & Sam O'Ferrell
date: June 06, 2024
format: html
---

## Library read in
```{r library read in}
library(tidyverse)
library(readr)
library(readxl)
```

## Data processing

### Step 1: Read in the data
```{r}
#reading in data from data folder
edu_data <- read_csv("./data/EDU01a.csv")

#reading specs of table to get column names
spec(edu_data)
```

### Step 2: Select only the Area_name (rename as area_name), STCOU and any columns that ends in “D”
```{r}
#saving edits as a new object
edu_edit <- edu_data |>
  #selecting only area_name, STCOU, and the columns that end with "D"
  select(Area_name, STCOU, ends_with("D")) |>
  rename("area_name" = "Area_name")
```

### Step 3: Converting to a long form 
```{r}
  #converting to a long form and moving the results to an "enrollment" column
edu_edit <- edu_edit |>
  pivot_longer(cols = 3:12, 
               values_to = "enrollment")
#printing to see the results
edu_edit
```

### Step 4: Converting year to numeric data and adding a column to represent measurement 
```{r}
#overwriting the object to save edits
edu_edit <- edu_edit |>
  #changing the values to reflect years using the data information sheet
  mutate( year = case_when(
    name == "EDU010187D" ~ 1987, 
    name == "EDU010188D" ~ 1988, 
    name == "EDU010189D" ~ 1989, 
    name == "EDU010190D" ~ 1990, 
    name == "EDU010191D" ~ 1991, 
    name == "EDU010192D" ~ 1992, 
    name == "EDU010193D" ~ 1993, 
    name == "EDU010194D" ~ 1994, 
    name == "EDU010195D" ~ 1995, 
    name == "EDU010196D" ~ 1996
  )) |>
  #creating a new column to represent which measurement grabbed
  mutate(measurement = substr(edu_edit$name, 1, 7))

#printing to see the results
edu_edit
```

### Step 5: Creating a county data set with state variable 
```{r}
#creating new data set with county level data 
edu_county <- subset(edu_edit, grepl(pattern = ", \\w\\w", edu_edit$area_name))

#overwriting the county object to save edits 
edu_county <- edu_county |>
  #adding a new column to represent county's state
  mutate(state = str_extract(edu_county$area_name, " \\w\\w")) |>
  #adjusting columns to look better
  select(area_name, state, STCOU, name, measurement, year, everything())

#changing the class of the county data set
class(edu_county) <- c("county", class(edu_county))

#printing the data set
edu_county
```

### Step 6: Creating a state data set with division variable 
```{r}
#creating new data set with state level data
edu_state <- subset(edu_edit, !(grepl(pattern = ", \\w\\w", edu_edit$area_name)))

#adjusting area_name values to easily identify the region
edu_state$area_name <- str_to_title(edu_state$area_name)

#adding new column to reflect division of the state
edu_state$division <- as.character(setNames(state.division, state.name)[edu_state$area_name])

#overwriting the state object to save edits 
edu_state <- edu_state |>
#fixing division data to reflect error condition and  
    mutate(division = case_when(
      
     str_to_title(area_name) %in% c("Connecticut", "Maine", "Massachusetts", 
                                    "New Hampshire", "Rhode Island", "Vermont") 
     ~ "New England",
     
    str_to_title(area_name) %in% c("New Jersey", "New York", "Pennsylvania") 
    ~ "Middle Atlantic",
    
    str_to_title(area_name) %in% c("Illinois", "Indiana", "Michigan", "Ohio", 
                                   "Wisconsin") 
    ~ "East North Central",
    
    str_to_title(area_name) %in% c("Iowa", "Kansas", "Minnesota", "Missouri", 
                                   "Nebraska", "North Dakota", "South Dakota") 
    ~ "West North Central",
    
    str_to_title(area_name) %in% c("Delaware", "Florida", "Georgia", "Maryland", 
                                   "North Carolina", "South Carolina", "Virginia", 
                                   "District of Columbia", "West Virginia")
    ~ "South Atlantic",
    
    str_to_title(area_name) %in% c("Alabama", "Kentucky", "Mississippi", "Tennessee") 
    ~ "East South Central",
    
    str_to_title(area_name) %in% c("Arkansas", "Louisiana", "Oklahoma", "Texas") 
    ~ "West South Central",
    
    str_to_title(area_name) %in% c("Arizona", "Colorado", "Idaho", "Montana",
                                   "Nevada", "New Mexico", "Utah", "Wyoming")
    ~ "Mountain",
    
    str_to_title(area_name) %in% c("Alaska", "California", "Hawaii", 
                                   "Oregon", "Washington") 
    ~ "Pacific",
    
    is.na(division) ~ "ERROR"
      )) |>
  #adjusting columns to look better
  select(area_name, division, STCOU, everything())

#printing the data set
edu_state
```

## Repeating with generic functions 

### Reading in the second data set:
```{r}
edu_data_b <- read_csv("./data/EDU01b.csv")

#reading specs of table to get column names
spec(edu_data_b)
```

### Repeating step 1 & 2 with option to specify name of value column 
```{r}
steps_1_and_2 <- function(data, value = "enrollment"){

# selecting the appropriate columns
  selected_columns <- data |>
       select(Area_name, STCOU, ends_with("D")) |>
       rename("area_name" = "Area_name") 
  
# pivoting the data, 'value' is defaulted to enrollment but can be changed
  pivoted_data <- selected_columns |> 
       pivot_longer(cols = 3:ncol(selected_columns), 
               values_to = value)
  
# returning the data 
  return(pivoted_data)
}
```

### Testing the function
```{r}
# the "value = " tells r what to label the column that is being pivoted
after_step_2 <- steps_1_and_2(data = edu_data_b, value = "enrollment")
after_step_2
```

### Writing function for step 3:
```{r}
step_3 <- function(data){
  
edit <- data |>
  #first we have to create a column corresponding to the last 2 digits
   mutate(last_2_digits =  substr(name,8,9),
  #then we will use logic to say if these are bigger than 24 the year must be in the 1900s
         year = as.numeric(ifelse(last_2_digits > 24, 
                       paste("19",last_2_digits, sep = ""), 
                       paste("20",last_2_digits, sep = ""))),
         measurement = substr(data$name, 1, 7)) |>
  select(-last_2_digits)
return(edit)
}
```

### Testing function for step 3:
```{r}
after_step_3 <- step_3(data = after_step_2)
after_step_3
```

### Writing function for step 5:
```{r}
#after already having the county data

step_5 <- function(data){
  with_state <- data |>
  mutate(state = substr(area_name, nchar(area_name) - 1,nchar(area_name))) |>
  #adjusting columns to look better
  select(area_name, state, STCOU, everything())  
  return(with_state)
}
```

### Testing step 5:
```{r}
#subsetting to work after step 4
county_practice <- subset(after_step_3, grepl(pattern = ", \\w\\w", after_step_3$area_name))

practice <- step_5(county_practice)
practice #works 
```

### Step 6 function (division):
```{r}
step_6_division <- function(data){
  title_case_data <- data |>
    # first we mutate to str_to_title and render that:
    mutate(area_name = str_to_title(area_name))

# fixing division data to reflect error condition 
  division_data <- title_case_data |>
    mutate(division = case_when(
      
     area_name %in% c("Connecticut", "Maine", "Massachusetts", 
                      "New Hampshire", "Rhode Island", "Vermont") 
     ~ "New England",
     
    area_name %in% c("New Jersey", "New York", "Pennsylvania") 
    ~ "Middle Atlantic",
    
    area_name %in% c("Illinois", "Indiana", "Michigan", "Ohio", "Wisconsin") 
    ~ "East North Central",
    
    area_name %in% c("Iowa", "Kansas", "Minnesota", "Missouri", "Nebraska", 
                     "North Dakota", "South Dakota") 
    ~ "West North Central",
    
    area_name %in% c("Delaware", "Florida", "Georgia", "Maryland", 
                     "North Carolina", "South Carolina", "Virginia", 
                     "District of Columbia", "West Virginia")
    ~ "South Atlantic",
    
    area_name %in% c("Alabama", "Kentucky", "Mississippi", "Tennessee") 
    ~ "East South Central",
    
    area_name %in% c("Arkansas", "Louisiana", "Oklahoma", "Texas") 
    ~ "West South Central",
    
    area_name %in% c("Arizona", "Colorado", "Idaho", "Montana", "Nevada", 
                     "New Mexico", "Utah", "Wyoming")
    ~ "Mountain",
    
    area_name %in% c("Alaska", "California", "Hawaii", "Oregon", "Washington") 
    ~ "Pacific",
    
  TRUE ~ "ERROR")) |>
  #adjusting columns to look better
  select(area_name, division, STCOU, everything())
  return(division_data)
}
```

### Testing function step 6:
```{r}
state <- subset(after_step_3, !(grepl(pattern = ", \\w\\w", after_step_3$area_name)))
practice_step_6 <- step_6_division(state)
practice_step_6
```

### Step 4 function: 
```{r}

step_4_5_6 <- function(data){
  #first we need to take the data and split it into 
  #two different data sets (county and state)
  county <- subset(data, grepl(pattern = ", \\w\\w", data$area_name))
  state <- subset(data, !(grepl(pattern = ", \\w\\w", data$area_name)))
  # now we apply both functions
  county_edit <- step_5(county)
  state_edit <- step_6_division(state)
  # return both data sets
  return(list("County_Data" = county_edit, "State_Data" = state_edit))
}
```

### Testing step 4, 5, 6 by using after step 3:
```{r}
final_data_practice <- step_4_5_6(after_step_3)
attributes(final_data_practice)
head(final_data_practice$County_Data)
head(final_data_practice$State_Data)

```

### Writing wrapper function
```{r}
my_wrapper <- function(url, default_var_name = "enrollment"){
# first we read in the csv
  result <- read_csv(url) |>
# then we apply steps 1 and 2
    steps_1_and_2(value = default_var_name)|>
# now we apply step 3
    step_3() |>
# finally steps 4,5,6 which returns a list of state and county data
    step_4_5_6()
return(result)
}
```

### Testing wrapper function
```{r}
# to use this function you provide a url and the default variable name
data_b_test <- my_wrapper(url = "./data/EDU01b.csv", default_var_name = "testing" )
data_b_test # lets gooooooo
```

### Applying the function to both data sets, resulting in two lists
```{r}
list_data_a <- my_wrapper(url = "./data/EDU01a.csv", default_var_name = "enrollment" )
list_data_b <- my_wrapper(url = "./data/EDU01b.csv", default_var_name = "enrollment" )
```

### Creating a function that combines the data sets
```{r}
combine <- function(data1, data2){
  # combine county data
  county_combined <- bind_rows(data1$County_Data, data2$County_Data)
  # combine state data
  state_combined <- bind_rows(data1$State_Data, data2$State_Data)
  # return list of combined data
  return(list("County_Data" = county_combined, 
              "State_Data" = state_combined))
}
```

applying function to both lists of data sets
```{r}
final_combined_data <- combine(list_data_a,list_data_b)
final_combined_data
```

## Writing generic functions for plotting the summary

### Plotting state data with enrollment mean and divison on y-axis and year on x-axis 
```{r}
#writing a function to plot based on state divisions 
plot.state <- function(data, var_name = "enrollment"){
  
  #creating a temporary object to store edits 
  data_summary <- data |>
    mutate(year = as.numeric(year)) |>
    
    #filtering out error values in division
    filter(division != "ERROR", !is.na(var_name)) |>
    
    #grouing by division and year 
    group_by(division, year) |>
    
    #summarizing the enrollment means
    summarise(enroll_mean = mean(get(var_name))) 
  
  #plotting the data as a line graph
  data_summary |> 
    
    ggplot(data_summary |> aes(x = year, y = enroll_mean, colour = division)) + 
    #setting the plot as line with additional aesthetic to make the lines look different 
    geom_line(aes(linetype = division)) + 
    #hiding the linetype legend
    guides(linetype = "none") +
    #changing the labels of x and y axis and the title
    labs(x = "Year", y = "Enrollment mean", title = "Enrollment changes over the years", colour = "Division") +
    #added to prevent scientific notion in the axis lables
    scale_y_continuous(labels = label_comma())
}

#testing the function 
plot.state(data = edu_state, var_name = "enrollment")
```

### Testing on final data set
```{r}
plot.state(final_combined_data$State_Data, var_name = "enrollment")
```

### County plot function with:
  -   data from the state specified or the default
  -   only the Area_names from the previous part
```{r}
plot.county <- function(data, state_name = "AL", interest = "top", 
                        x = "5", var_name = "enrollment"){
 # first create data set that has areas we are interested in
   if (interest == "top"){
  summary <- data |>
  filter(state == state_name) |>
  group_by(area_name) |>
  summarise(enroll_mean = mean(get(var_name))) |>
  arrange(desc(enroll_mean)) |>
  slice(1:x) |>
  select(area_name)
  }
  else if(interest == "bottom"){
  summary <- data |>
  filter(state == state_name) |>
  group_by(area_name) |>
  summarise(enroll_mean = mean(get(var_name))) |>
  arrange(enroll_mean) |>
  slice(1:x) |>
  select(area_name)
  }
# now combine this with original data set to plot values:
  new <- semi_join(data, summary, by = "area_name")
  ggplot(new |> mutate(year = as.numeric(year)), aes(x = year, y = get(var_name), colour = area_name)) + 
    theme(legend.position="bottom") +
    geom_line(aes(linetype = area_name)) + 
    guides(linetype = "none") +
    labs(x = "Year", y = "Enrollment", title = "Enrollment changes over the years", colour = "County") +
    #added to prevent scientific notion in the axis lables
    scale_y_continuous(labels = label_comma())
  
}
```

### Testing the county plot function
```{r}
plot.county(state = "NC", 
            data = final_combined_data$County_Data, 
            var_name = "enrollment", 
            interest = "top", 
            x = 15)
```

## Final Report 

### Run our data processing function on two enrollment URLs:
```{r}
list_data_a <- my_wrapper(
  url = "https://www4.stat.ncsu.edu/~online/datasets/EDU01b.csv", 
  default_var_name = "enrollment" )
list_data_b <- my_wrapper(
  url = "https://www4.stat.ncsu.edu/~online/datasets/EDU01a.csv", 
  default_var_name = "enrollment" )
```

### Run our data combining function to put these into one object (with two data frames):
```{r}
final_combined_data <- combine(list_data_a,list_data_b)
final_combined_data
```

### Use the plot function on the state data frame:
```{r}
plot.state(data = final_combined_data$State_Data, 
           var_name = "enrollment")
```


### Use the plot function on the county data frame:
  -   Specifying the state to be “NC”, the group being the top, the number looked at being 20
```{r}
plot.county(state = "NC", 
            data = final_combined_data$County_Data, 
            interest = "top", 
            x = 20)
```
– Specifying the state to be “SC”, the group being the bottom, the number looked at being 7
```{r}
plot.county(state = "SC", 
            data = final_combined_data$County_Data, 
            interest = "bottom", 
            x = 7)
```
– Once without specifying anything (defaults used)
```{r}
plot.county(data = final_combined_data$County_Data )
```
– Once specifying the state to be “PA”, the group being the top, the number looked at being 8
```{r}
plot.county(state = "PA", 
            data = final_combined_data$County_Data, 
            var_name = "enrollment", 
            interest = "top", 
            x = 8)
```

### Reading similar data sets and applying our functions:
```{r}
list_data_PST01a <- my_wrapper(
  url = "https://www4.stat.ncsu.edu/~online/datasets/PST01a.csv", 
  default_var_name = "enrollment" )
list_data_PST01b <- my_wrapper(
  url = "https://www4.stat.ncsu.edu/~online/datasets/PST01b.csv", 
  default_var_name = "enrollment" )
list_data_PST01c <- my_wrapper(
  url = "https://www4.stat.ncsu.edu/~online/datasets/PST01c.csv", 
  default_var_name = "enrollment" )
list_data_PST01d <- my_wrapper(
  url = "https://www4.stat.ncsu.edu/~online/datasets/PST01d.csv", 
  default_var_name = "enrollment" )
```

### Combine all 4 data sets:
```{r}
pre_final_combined_data_ab <- combine(list_data_PST01a,list_data_PST01b)
pre_final_combined_data_cd <- combine(list_data_PST01c,list_data_PST01d)
final_combined_data_PST <- combine(pre_final_combined_data_ab,pre_final_combined_data_cd)
```

